Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-03-17 paper.qmd
Removed prediction code

```{r}
#| label: dist-plot

df_bin_two_curvy_temp <- df_bin_two_curvy
```

@@

```{r}
#| label: model-prediction-umap-quollr
#| eval: false

true_pred_df <- predict_emb(
  test_data = true_model_df,
  df_bin_centroids = df_bin_centroids_two_curvy,
  df_bin = df_bin_two_curvy_temp,
  type_NLDR = "UMAP"
)

# Compute radius r
a1 <- calc_bins_y(
  bin1 = 10, 
  r2 = r2, 
  q = 0.1)$a1

r <- a1/2

# Function to jitter points within a circumcircle of a hexagon
jitter_within_circumcircle <- function(center_x, center_y, radius, num_points) {
  theta <- runif(num_points, 0, 2 * pi)
  r <- radius * sqrt(runif(num_points))
  x <- center_x + r * cos(theta)
  y <- center_y + r * sin(theta)
  jittered_points <- data.frame(UMAP1 = x, UMAP2 = y, ID = points_in_hex$ID)
  return(jittered_points)
}

# Jittered points data frame
jittered_points_df <- data.frame()

# Iterate through each hexagon and jitter points within the circumcircle
for (hex_id in unique(true_pred_df$pred_hb_id)) {
  # hex_points <- hex_grid_nonempty %>% filter(hex_poly_id == hex_id)
  # center_x <- mean(hex_points$x)
  # center_y <- mean(hex_points$y)
  # points_in_hex <- true_pred_df %>% filter(pred_UMAP_1 >= min(hex_points$x) & pred_UMAP_1 <= max(hex_points$x) & pred_UMAP_2 >= min(hex_points$y) & pred_UMAP_2 <= max(hex_points$y))

  points_in_hex <- true_pred_df |>
    filter(pred_hb_id == hex_id)

  center_x <- points_in_hex |> 
    pull(pred_UMAP_1) |> 
    unique()

  center_y <- points_in_hex |> 
    pull(pred_UMAP_2) |> 
    unique()

  if (nrow(points_in_hex) > 0) {
    jittered_points <- jitter_within_circumcircle(center_x, center_y, r, nrow(points_in_hex))
    jittered_points_df <- rbind(jittered_points_df, jittered_points)
  }
}

plot_predict_tsne_two_curvy <- ggplot(
  data = true_pred_df, 
  aes(x = pred_UMAP_1,
      y = pred_UMAP_2)) +
  geom_point(alpha = 0.5) +
  interior_annotation("c", c(0.08, 0.93), cex = 1.5)
```

@@

error_two_non_linear_diff_shaped_close_clusters_umap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_tsne.rds")

## Find the minimum MSE when have duplicate a1
error_two_non_linear_diff_shaped_close_clusters_umap <- error_two_non_linear_diff_shaped_close_clusters_umap |>
  group_by(a1) |>
  filter(bin1 == min(bin1)) |>
  ungroup() |>
  mutate(prop_dens = 1/(b*side_length^2))

base_line_prop_dens <- error_two_non_linear_diff_shaped_close_clusters_umap |>
  filter(a1 == min(a1)) |>
  pull(prop_dens)

error_two_non_linear_diff_shaped_close_clusters_umap <- error_two_non_linear_diff_shaped_close_clusters_umap |>
  mutate(prop_comp = prop_dens/base_line_prop_dens)

mse_two_curvy_b <- ggplot(error_two_non_linear_diff_shaped_close_clusters_umap, 
                     aes(x = a1, 
                         y = MSE)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = sort(unique(round(error_two_non_linear_diff_shaped_close_clusters_umap$a1, 2)))[c(2, 5, 8, 10, 12, 14, 15, 16, 17)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "MSE") +
  ggtitle("(a)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))

prop_dens_a1 <- ggplot(error_two_non_linear_diff_shaped_close_clusters_umap,
                     aes(x = a1,
                         y = prop_comp)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_non_linear_diff_shaped_close_clusters_umap$a1, 2)))[c(2, 5, 8, 10, 12, 14, 15, 16, 17)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = paste("Relative proportion density")) +
  ggtitle("(d)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
axis.text = element_text(size = 12))

error_two_non_linear_diff_shaped_close_clusters_umap <- error_two_non_linear_diff_shaped_close_clusters_umap |>
  mutate(prop_bins = b_non_empty/b)

a1_m_two_curvy <- ggplot(error_two_non_linear_diff_shaped_close_clusters_umap,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_non_linear_diff_shaped_close_clusters_umap$a1, 2)))[c(2, 5, 8, 10, 12, 14, 15, 16, 17)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression(paste("proportion of non-empty bins ", bgroup("(", frac(m, b), ")")))) +
  ggtitle("(b)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
axis.text = element_text(size = 12))

@@

error_two_non_linear_diff_shaped_close_clusters_umap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_umap.rds")
error_two_non_linear_diff_shaped_close_clusters_tsne <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_tsne.rds")
error_two_non_linear_diff_shaped_close_clusters_phate <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_phate.rds")
error_two_non_linear_diff_shaped_close_clusters_trimap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_trimap.rds")
error_two_non_linear_diff_shaped_close_clusters_pacmap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_pacmap.rds")
error_two_non_linear_diff_shaped_close_clusters_tsne2 <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_tsne2.rds")

@@

error_two_non_linear_diff_shaped_close_clusters <- bind_rows(error_two_non_linear_diff_shaped_close_clusters_umap, 
                         error_two_non_linear_diff_shaped_close_clusters_tsne,
                         error_two_non_linear_diff_shaped_close_clusters_phate,
                         error_two_non_linear_diff_shaped_close_clusters_trimap,
                         error_two_non_linear_diff_shaped_close_clusters_pacmap,
                         error_two_non_linear_diff_shaped_close_clusters_tsne2)

>>>>

@@

@@

error_two_curvy_umap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_tsne.rds")

## Find the minimum MSE when have duplicate a1
error_two_curvy_umap <- error_two_curvy_umap |>
  group_by(a1) |>
  filter(bin1 == min(bin1)) |>
  ungroup() |>
  mutate(prop_dens = 1/(b*side_length^2))

base_line_prop_dens <- error_two_curvy_umap |>
  filter(a1 == min(a1)) |>
  pull(prop_dens)

error_two_curvy_umap <- error_two_curvy_umap |>
  mutate(prop_comp = prop_dens/base_line_prop_dens)

mse_two_curvy_b <- ggplot(error_two_curvy_umap, 
                     aes(x = a1, 
                         y = MSE)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(2, 5, 8, 10, 12, 14, 15, 16, 17)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "MSE") +
  ggtitle("(a)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))

prop_dens_a1 <- ggplot(error_two_curvy_umap,
                     aes(x = a1,
                         y = prop_comp)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(2, 5, 8, 10, 12, 14, 15, 16, 17)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = paste("Relative proportion density")) +
  ggtitle("(d)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
axis.text = element_text(size = 12))

error_two_curvy_umap <- error_two_curvy_umap |>
  mutate(prop_bins = b_non_empty/b)

a1_m_two_curvy <- ggplot(error_two_curvy_umap,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(2, 5, 8, 10, 12, 14, 15, 16, 17)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression(paste("proportion of non-empty bins ", bgroup("(", frac(m, b), ")")))) +
  ggtitle("(b)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
axis.text = element_text(size = 12))

@@

error_two_curvy_umap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_umap.rds")
error_two_non_linear_diff_shaped_close_clusters_tsne <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_tsne.rds")
error_two_non_linear_diff_shaped_close_clusters_phate <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_phate.rds")
error_two_non_linear_diff_shaped_close_clusters_trimap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_trimap.rds")
error_two_non_linear_diff_shaped_close_clusters_pacmap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_pacmap.rds")
error_two_non_linear_diff_shaped_close_clusters_tsne2 <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_tsne2.rds")

@@

error_two_non_linear_diff_shaped_close_clusters <- bind_rows(error_two_curvy_umap, 
                         error_two_non_linear_diff_shaped_close_clusters_tsne,
                         error_two_non_linear_diff_shaped_close_clusters_phate,
                         error_two_non_linear_diff_shaped_close_clusters_trimap,
                         error_two_non_linear_diff_shaped_close_clusters_pacmap,
                         error_two_non_linear_diff_shaped_close_clusters_tsne2)
